---
title: "FeCit_Ymaze_pilot"
author: "Karissa Barthelson"
date: "2022-10-31"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  autodep = TRUE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center", 
  fig.retina = 2,
  out.width ="100%", 
  out.height = "100%"
)
```

```{r}
library(tidyverse)
library(readxl)
library(here)
library(magrittr)
library(scales)
library(readxl)
library(ggpubr)
library(ggeasy)
library(ggfortify)
library(ggbeeswarm)
library(ggforce)
library(ggrepel)
library(kableExtra)

# stat analysis
library(broom)
library(lme4)
library(performance)
library(car)
library(emmeans)
library(glmmTMB)
library(MASS)

# set the default theme for ggplot as theme_bw
theme_set(theme_bw())
```

```{r}

# read in the processed data and metadata
meta <- read_xlsx("data/2022Oct19 ip injection FeCitrate naglu hom x het 6m meta.xlsx") %>% 
  mutate(fish_id = as.character(fish_id), 
         treatment = factor(treatment, 
                            levels = c("0.85% saline", 
                                       "10 µg Fe-citrate", 
                                       "100 µg Fe-citrate"))
  )

final_data <- read_csv("data/ymaze/processed_data/fecit/final_output.csv") %>% 
  dplyr::select(-1) %>% 
  mutate(fish_id = factor(fish_id)) %>% 
  left_join(meta) %>% 
  dplyr::filter(genotype%in% c("het", "hom"))
#view(final_data)

# make an object which converts the final data to long format. for easier plotting in ggplot
final_data_long <- final_data %>%
  gather(key = "tetras", value = "Count", # convert to long format
         grep("[L|R]{4}", 
              colnames(.))
         )

# also make an object which sums the tetragrams over the hour
final_data_summedoverbins <- final_data %>%
  gather(key = "tetras", value = "Count", # convert to long format
         grep("[L|R]{4}", # select the columns which contain a L or a R four times
              colnames(.))
         ) %>% 
  group_by(fish_id, tetras) %>% 
  mutate(x = sum(Count)) %>% # sum the tetragram counts per fish_id
  dplyr::select(colnames(meta), tetras, x) %>% 
  unique
```

# Introduction

In this analysis, I will analyse behavioural data generated for zebrafish arising from a lay from  *naglu* A603fs hom x het after treatment of different amounts of ferric citrate in the Zantiks LT Y-maze. 

Note that these fish were tested while construction was occuring on the floor below us. There was often loud noises occuring. 

Fish were isolated for 45 mins, then placed in the mazes alone for 1 hour. Raw data was collected, and metadata was recorded (i.e. sex, time, etc). These raw data spreadsheets were batch processed in the BatchProcess R script in the `code` folder to produce the `final_data` object. This object contains the tetragram frequencies (see below). Note that fish were genotyped after data collection so that we are blinded from any observer bias until after raw data collection. Additionally, the videos were manually observed to ensure proper tracking. Any fish where the tracking was not perfect was omitted. 

```{r}
 read_csv("data/ymaze/raw_data/fecit/ymaze_inLT-20221027T103409.csv", skip =4, col_names = F) %>% 
  head %>% 
  dplyr::select(-X2, -X3) %>% 
  set_colnames(c("Time", "fish",'enter or exit', "zone")) %>% 
  kable(caption = "example of raw data") %>% 
  kable_styling(full_width = F)
```


```{r}
final_data %>% 
  head %>% 
  kable(caption = "Example of the processed data") %>% 
  kable_styling() 
```


```{r}
# define fish which had the tracking issues
fish2omit <- final_data_summedoverbins %>% 
  dplyr::filter(x > 200) %>% 
  .$fish_id %>% 
  unique

# remove them
final_data %<>%
  dplyr::filter(!(fish_id %in%fish2omit))

final_data_long%<>%
  dplyr::filter(!(fish_id %in% fish2omit))

final_data_summedoverbins%<>%
  dplyr::filter(!(fish_id %in% fish2omit))
```

# check proportions of fish 

After omitting the fish with tracking issues, only 3 het fish treated with 0.85% saline remain. This is not great for the stats, but this is only a pilot overall. 

```{r}
final_data %>% 
  dplyr::select(colnames(meta)) %>% 
  dplyr::distinct() %>% 
  group_by(genotype, sex, treatment, dpi) %>% 
  mutate(n = n()) %>% 
  ggplot(aes(x = genotype)) +
  geom_bar(aes(fill = treatment), 
           position = "dodge", 
           colour = "black") +
  facet_wrap(~dpi, nrow = 1) +
  scale_fill_viridis_d() + 
  scale_y_continuous(breaks = seq(1:9)) +
  theme(legend.position = "bottom"
        )+
  labs(title = "Number of fish per genotype, sex, treatment and dpi in the qPCR analysis",
       y = "Number of fish", 
       x = "")
```


# Visualisation of raw data

Cleal et al. showed that zebrafish naturally perform more of the alternation tetragrams (LRLR and RLRL) in a Y-maze. We also see thgis somewhat here. 

```{r}
final_data_summedoverbins %>%
  dplyr::distinct(x, .keep_all = T) %>%
  ggplot(aes(x = tetras, y = x)) +
  geom_jitter(aes(colour = tetras,
                  shape = sex)) +
  geom_boxplot(outlier.shape = NA,
               # aes(fill = Genotype)
               fill = NA
               ) +
  scale_fill_viridis_d() +
  labs(y = "Number of tetragrams",
       colour = "Tetragram",
       x = "Tetragram")+
  theme(legend.position = "bottom")  +
  ggtitle("Total number of 16 possible tetragrams performed by zebrafish in a Y-maze\nduring a 1 hour search period") 
```

We can also overlay the genotype boxplots as shown below. Looks like the heterozyous and homozygous fish perform more alternations than the WT fish. I will look into this more later in the analysis

```{r}
final_data_summedoverbins %>%
  dplyr::distinct(x, .keep_all = T) %>%
  ggplot(aes(x = tetras, y = x)) +
  geom_jitter(aes(colour = tetras,
                  shape = sex)) +
  geom_boxplot(outlier.shape = NA,
               aes(fill = genotype),
               alpha = 0.5,
               ) +
  scale_fill_viridis_d() +
  labs(y = "Number of tetragrams",
       colour = "Tetragram",
       x = "Tetragram")+
  theme(legend.position = "bottom")  +
  facet_wrap(~treatment, ncol = 1) +
  easy_rotate_x_labels(angle = -45) +
  annotate("rect", # add some boxes aeround the alts 
           xmin = 5.5, xmax = 6.5, 
           ymin = -1, ymax = 200, 
             alpha = 0, color= "red") +
  annotate("rect", 
           xmin = 10.5, xmax = 11.5, 
           ymin = -1, ymax = 200,
           alpha = 0, color= "red") +
  ggtitle("Total number of 16 possible tetragrams performed by zebrafish in a Y-maze\nduring a 1 hour search period") 
```

The alternation tetragrams are the tetragram of interest (the measure of working memory). Below indicates the number of tetragrams performed by zerbafish across the 6 x 10 min blocks of the hour they spent in the maze.

```{r}
final_data_long %>%
  dplyr::distinct(rel_alts, .keep_all = T) %>%
  mutate(binforvis = case_when(
    bin == 1 ~ "0-10 mins", 
    bin == 2 ~ "10-20 mins", 
    bin == 3 ~ "20-30 mins", 
    bin == 4 ~ "30-40 mins", 
    bin == 5 ~ "40-50 mins", 
    bin == 6 ~ "50-60 mins"
    )) %>% 
  ggplot(aes(x = treatment, y = alts)) +
  geom_boxplot(outlier.shape = NA,
               aes(fill = genotype), 
               alpha = 0.5) +
  geom_point(aes(colour = genotype), 
             position = position_jitterdodge()) +
  facet_wrap(~binforvis, nrow = 1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), 
        legend.position = "bottom") +
  ylab("Frequency of alternation tetragrams (LRLR + RLRL)") +
  scale_colour_viridis_d(end = 0.75) +
  scale_fill_viridis_d(end = 0.75) 
   # geom_label_repel(aes(label = fish_id),
   #                 data = . %>%
   #                   dplyr::filter(fish_id == 1))
```

We can also display this as the *relative* amount of alternation tetragrams performed by zebrafish in a 1 hour search period. This will control for how active each fish is. Again, fish 18 and 24 looks like outliers.

```{r}
final_data_long %>%
  dplyr::distinct(rel_alts, .keep_all = T) %>%
  mutate(binforvis = case_when(
    bin == 1 ~ "0-10 mins", 
    bin == 2 ~ "10-20 mins", 
    bin == 3 ~ "20-30 mins", 
    bin == 4 ~ "30-40 mins", 
    bin == 5 ~ "40-50 mins", 
    bin == 6 ~ "50-60 mins"
    )) %>% 
  ggplot(aes(x = treatment, y = rel_alts)) +
  geom_boxplot(outlier.shape = NA,
               aes(fill = genotype), 
               alpha = 0.5) +
  geom_point(aes(colour = genotype), 
             position = position_jitterdodge()) +
  facet_wrap(~binforvis, nrow = 1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), 
        legend.position = "bottom") +
  ylab("Relative number of alternation tetragrams\n(LRLR + RLRL) / total_turns") +
  scale_colour_viridis_d(end = 0.75) +
  scale_fill_viridis_d(end = 0.75) 
```

# Locomotor defect

Alteration to locomotor activity could be a confounding effect with genotype to alternation (LRLR + RLRL) frequency. If mutant and/or treated fish had some form of locomotor change, they could be performing more or less alternation tetragrams not due to impairment of spatial working memory. To account for this, I will test whether the genotype and/or treatment of the fish had a significant effect on:

* 1) the total number of turns the fish performed in the hour spent in the maze or 
* 2) the average time spent in each zone of the maze.


## total number of turns

I first visualised the total number of turns . Some differences are obsevred between the hets and the homs. Particularly in the 100 µg Fe-citrate treatment group.

```{r}
final_data_summedoverbins %>%
  ungroup() %>%
  group_by(fish_id) %>%
  mutate(total_turns = sum(x)) %>%
  dplyr::select(colnames(meta), total_turns) %>%
  unique %>%
  ggplot(aes(x = genotype,y = total_turns)) +
  geom_violin(aes(fill = genotype),
               alpha = 0.5) +
  geom_boxplot(aes(colour = genotype),
               fill = NA,
               width = 0.25,
               colour= "black") +
  geom_quasirandom(aes(shape = sex), 
                   size = 2) +
  scale_fill_viridis_d(option = "viridis") +
  scale_colour_viridis_d(option = "viridis") +
  facet_wrap(~treatment) +
  labs(y = "Total number of turns", 
       title = "Total number of turns performed by fish in the Y-maze in 1 hour", 
       subtitle = "According to treatment")
```

The plot below is grouped by genotype rather than treartment. Not much difference between treatment groups within genotypes. 

```{r}
final_data_summedoverbins %>%
  ungroup() %>%
  group_by(fish_id) %>%
  mutate(total_turns = sum(x)) %>%
  dplyr::select(colnames(meta), total_turns) %>%
  unique %>%
  ggplot(aes(x = treatment,y = total_turns)) +
  geom_violin(aes(fill = genotype),
               alpha = 0.5) +
  geom_boxplot(aes(colour = genotype),
               fill = NA,
               width = 0.25,
               colour= "black") +
  geom_quasirandom(aes(shape = sex), 
                   size = 2) +
  scale_fill_viridis_d(option = "viridis") +
  scale_colour_viridis_d(option = "viridis") +
  facet_wrap(~genotype) +
  labs(y = "Total number of turns", 
       title = "Total number of turns performed by fish in the Y-maze in 1 hour", 
       subtitle = "According to treatment")
```

I also want to look at whether there is a batch effect of test. There were a total of 6 tests (start times indicated in the plot) performed in one of two zantiks units (139 and 146). Each test contained 8 fish which were randomly assigned. A bit of fvariation is observed between start times, but this would be expected considering they are performed throughoutthe day, and have had sligtly different environmental conditions (e.g. ,vibrations from the construction, people talking out in the hall etc), as well as circadian differences. 

```{r}
final_data_summedoverbins %>%
  ungroup() %>%
  group_by(fish_id) %>%
  mutate(total_turns = sum(x)) %>%
  dplyr::select(colnames(meta), total_turns) %>%
  unique %>%
  mutate(`start time` = factor(`start time`, levels = c(
    "1034am", 
    "1048am",
    "1152am",
    "1156am",
    "105pm",
    "110pm"
  ))
  ) %>% 
  ggplot(aes(x = `start time`, y = total_turns)) +
  geom_boxplot(aes(fill = `start time`),
               outlier.shape = NA,
               width = 0.25,
               alpha = 0.5,
               colour= "black") +
  geom_jitter(aes(colour = `start time`),
              size = 3) +
  labs(y = "Total number of turns", 
       title = "Total number of turns performed by fish in the Y-maze in 1 hour", 
       subtitle = "According to start time of day of the test")
```

I also had a look at whether the genotype and treatment groups could explain some of the apparent variation within tests. This does seem to be the case as observed in the fig below. the genotypes can sometimes show differences. 

```{r}
final_data_summedoverbins %>%
  ungroup() %>%
  group_by(fish_id) %>%
  mutate(total_turns = sum(x),
         `start time` = factor(`start time`, levels = c(
           "1034am", 
           "1048am",
           "1152am",
           "1156am",
           "105pm",
           "110pm"
         ))) %>%
  dplyr::select(colnames(meta), total_turns) %>%
  unique %>%
  ggplot(aes(x = genotype, y = total_turns)) +
  geom_violin(aes(fill = genotype),
               alpha = 0.5) +
  geom_boxplot(aes(colour = genotype),
               fill = NA,
               width = 0.25,
               colour= "black") +
  geom_jitter(aes(shape = treatment), 
                   size = 3) +
  scale_fill_viridis_d(option = "viridis") +
  scale_colour_viridis_d(option = "viridis") +
  facet_wrap(~ymazeUni+`start time`) +
  labs(y = "Total number of turns", 
       title = "Total number of turns performed by fish in the Y-maze in 1 hour", 
       subtitle = "According to start time of day of the test")
```

Trying to determine whether the effect of treatment group and start time is more difficult to say, as there is often < 2 fish per treatment group within a particular test. This is just due to the nature of performing this pilot experiment and will not be such as issue next time (when we only have 2 treatment groups, saline and iron)

```{r}
final_data_summedoverbins %>%
  ungroup() %>%
  group_by(fish_id) %>%
  mutate(total_turns = sum(x)) %>%
  dplyr::select(colnames(meta), total_turns) %>%
  unique %>%
  ggplot(aes(x = treatment, y = total_turns)) +
  geom_violin(aes(fill = treatment),
               alpha = 0.5) +
  geom_boxplot(aes(colour = treatment),
               fill = NA,
               width = 0.25,
               colour= "black") +
  geom_jitter(aes(shape = genotype), 
                   size = 3) +
  scale_fill_viridis_d(option = "viridis") +
  scale_colour_viridis_d(option = "viridis") +
  facet_wrap(~ymazeUni+`start time`) +
  easy_rotate_x_labels(angle = -45) +
  labs(y = "Total number of turns", 
       title = "Total number of turns performed by fish in the Y-maze in 1 hour", 
       subtitle = "According to start time of day of the test")
```

## statistical test

To test whether the genotype and/or treatment have a significant effect overall on the total number of turns, I fitted the summed over bins data to a negative binomial generalised linear mixed effect model (negative binomial). No statistical evidence is observed here. The closest is the treatment 100 µg Fe-citrate group, which makes sense considering the plots above show the most difference. Since the start time of the test doesnt have a logical pattern (i.e. more active in morning), I've included it in the model as a random effect. 


```{r}
final_data_summedoverbins %>%
  ungroup() %>%
  group_by(fish_id) %>%
  mutate(total_turns = sum(x)) %>%
  dplyr::select(colnames(meta), total_turns) %>%
  unique %>%
  glmer.nb(total_turns ~ genotype*treatment + (1|`start time`),
         data = .) %>%
  summary %>%
  .$coef %>%
  as.data.frame() %>% 
  dplyr::rename(pval = "Pr(>|z|)") %>% 
  kable(caption = "Negative binomial GLM model coefficients.\n
      None of the factors have a significant effect on the total number of turns" ) %>%
  kable_styling() %>% 
  row_spec(row = c(5,6), background = "#ccffef")
```

## turns across the hour. 

I also want to look at whether the fish display changes to activity across the hour (i.e. are they more active when they are first placed in the maze?). 

```{r}
final_data_long %>%
  dplyr::distinct(rel_alts, .keep_all = T) %>%
  mutate(binforvis = case_when(
    bin == 1 ~ "0-10 mins", 
    bin == 2 ~ "10-20 mins", 
    bin == 3 ~ "20-30 mins", 
    bin == 4 ~ "30-40 mins", 
    bin == 5 ~ "40-50 mins", 
    bin == 6 ~ "50-60 mins"
    )) %>% 
  ggplot(aes(x = treatment, y = total_turns)) +
  geom_boxplot(aes(fill = genotype),
           position = "dodge",
           alpha = 0.5,
           outlier.shape = NA) +
   geom_point(aes(colour = genotype), 
             position = position_jitterdodge()) +
  facet_wrap(~binforvis, nrow = 1)+
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1, vjust = 1), 
        legend.position = "bottom") +
  labs(title = "Number of turns performed by zebrafish across an hour ", 
       subtitle = "Accoring to treatment and genotype",
       y = "Number of turns") +
  scale_colour_viridis_d(end = 0.75) +
  scale_fill_viridis_d(end = 0.75)
```

Looking at the changes to the total number of turns across the bins within genoytpes. The overall distibution looks simialr in each mutants, with the activity decreasing as time goes on. 

```{r}
final_data_long %>%
  dplyr::distinct(rel_alts, .keep_all = T) %>%
  mutate(binforvis = case_when(
    bin == 1 ~ "0-10 mins", 
    bin == 2 ~ "10-20 mins", 
    bin == 3 ~ "20-30 mins", 
    bin == 4 ~ "30-40 mins", 
    bin == 5 ~ "40-50 mins", 
    bin == 6 ~ "50-60 mins"
    )) %>% 
  ggplot(aes(x = binforvis, y = total_turns)) +
  geom_boxplot(aes(fill = treatment),
               position = "dodge",
               alpha = 0.5,
               outlier.shape = NA) +
  geom_point(aes(colour = treatment), 
             position = position_jitterdodge()) +
  facet_wrap(~genotype, nrow = 1, scales = "free_x")+
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1, vjust = 1)) +
  labs(title = "Number of turns performed by zebrafish across an hour ", 
       subtitle = "Accoring to treatment and genotype",
       y = "Number of turns") +
  scale_colour_viridis_d(end = 0.75) +
  scale_fill_viridis_d(end = 0.75)
```

Finally, want to vis within treatments

```{r}
final_data_long %>%
  dplyr::distinct(rel_alts, .keep_all = T) %>%
  mutate(binforvis = case_when(
    bin == 1 ~ "0-10 mins", 
    bin == 2 ~ "10-20 mins", 
    bin == 3 ~ "20-30 mins", 
    bin == 4 ~ "30-40 mins", 
    bin == 5 ~ "40-50 mins", 
    bin == 6 ~ "50-60 mins"
    )) %>% 
  ggplot(aes(x = binforvis, y = total_turns)) +
  geom_boxplot(aes(fill = genotype),
               position = "dodge",
               alpha = 0.5,
               outlier.shape = NA) +
  geom_point(aes(colour = genotype), 
             position = position_jitterdodge()) +
  facet_wrap(~treatment, ncol = 1)+
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1, vjust = 1)) +
  labs(title = "Number of turns performed by zebrafish across an hour ", 
       subtitle = "Accoring to treatment and genotype",
       y = "Number of turns") +
  scale_colour_viridis_d(end = 0.75) +
  scale_fill_viridis_d(end = 0.75)
```

## statistical test

To test whether the genotype and/or treatment have a significant effect overall on the total number of turns, I fitted the per bin data to a negative binomial generalised linear mixed effect model. 
Since the start time of the test doesnt have a logical pattern (i.e. mrore active in morning), I've included it in the model as a random effect. 

The bin has a highly signifdicant effect on the total turns. This is consistent with what we have seen in the past using this test. The fish are more active when they first enter the maze and it drops off as they become more accustomed to it. Some statistical evidence is obsered for the `genotypehom:treatment10 µg Fe-citrate`effect. However, I wonder is this is truly real, as we only had 3 hets. 

```{r}
final_data_long %>%
  dplyr::distinct(rel_alts, .keep_all = T) %>%
  mutate(binforvis = case_when(
    bin == 1 ~ "0-10 mins", 
    bin == 2 ~ "10-20 mins", 
    bin == 3 ~ "20-30 mins", 
    bin == 4 ~ "30-40 mins", 
    bin == 5 ~ "40-50 mins", 
    bin == 6 ~ "50-60 mins"
    )) %>% 
  glmer.nb(total_turns ~ genotype*treatment + binforvis + (1|`start time`),
         data = .) %>%
  summary %>%
  .$coef %>%
  as.data.frame() %>% 
  dplyr::rename(pval = "Pr(>|z|)") %>% 
  kable(caption = "Negative binomial GLM model coefficients.\n
      None of the factors have a significant effect on the total number of turns" ) %>%
  kable_styling() %>% 
  row_spec(row = c(10,11), background = "#ccffef")
```

# time spent in each zone of the Y-maze

I next will assess whether fish in each maze spend more or less time in each zone of the Y-maze (i.e. in each arm of the maze or the middle). 

In the plot below, the fish appear to spend less time in zone 4. This can be explained as zone 4 is the middle zone and the fish generally swim straight through it.  

```{r}
read_csv("data/ymaze/processed_data/fecit/time_in_zone.csv") %>%
  mutate(fish_id = as.character(fish_id)) %>%
  dplyr::select(fish_id, bin, zone, time_in_zone) %>%
  left_join(meta) %>%
  dplyr::filter(genotype %in% c("het", "hom")) %>% 
  group_by(fish_id, zone) %>%
  mutate(total_timeInZone = sum(time_in_zone), 
         zone = paste0("zone ", zone)) %>%
  dplyr::distinct(total_timeInZone, .keep_all = TRUE) %>%
  ggplot(aes(x = treatment, y = total_timeInZone/60)) +
  geom_quasirandom(aes(colour = genotype)
                  ) +
  geom_boxplot(aes(fill = genotype),
               outlier.shape = NA,
                alpha = 0.5) +
  facet_wrap(~zone, nrow = 1) +
  scale_y_log10() +
  scale_fill_viridis_d(end = 0.75) +
  scale_colour_viridis_d(end = 0.75) +
  easy_rotate_x_labels(angle = -45) +
  labs(y = "Total time spent in each zone (mins, log scale)", 
       title = "Total time spent in each zone")
```

I was curious to see whether which maze the fish were in (i.e. top left, middle etc) would have an effect on whether they spend more or less time in one of the arms of the mazes. There are 8 ymazes in each zantiks unit. inspection of the total time spent in each zone of the maze across the 8 possible positions in the zantiks unit didnt reveal any trends. 

There might have been something in position 6, zone 4. But this is only 2 data points and so not really reliable

```{r}
read_csv("data/ymaze/processed_data/fecit/time_in_zone.csv") %>%
  mutate(fish_id = as.character(fish_id)) %>%
  dplyr::select(fish_id, bin, zone, time_in_zone) %>%
  left_join(meta) %>%
  dplyr::filter(genotype %in% c("het", "hom")) %>% 
  group_by(fish_id, zone) %>%
  mutate(total_timeInZone = sum(time_in_zone), 
         zone = paste0("zone ", zone)) %>%
  dplyr::distinct(total_timeInZone, .keep_all = TRUE) %>%
  ggplot(aes(x = zone, y = total_timeInZone/60)) +
  geom_quasirandom(aes(colour = genotype, 
                       shape = treatment), 
                   size = 2
                  ) +
  geom_boxplot(aes(fill = genotype),
               outlier.shape = NA,
                alpha = 0.5) +
  facet_wrap(~YmazePosition) +
  scale_y_log10() +
  scale_fill_viridis_d(end = 0.75) +
  scale_colour_viridis_d(end = 0.75) +
  easy_rotate_x_labels(angle = -45) +
  labs(y = "Total time spent in each zone (mins, log scale)", 
       title = "Total time spent in each zone", 
       subtitle = "accoring to position in the ymaze")
```

I alslo wanted to look at the average time spent in each zone. 

```{r}
read_csv("data/ymaze/processed_data/fecit/time_in_zone.csv") %>%
  mutate(fish_id = as.character(fish_id)) %>%
  dplyr::select(fish_id, bin, zone, time_in_zone) %>%
  left_join(meta) %>%
  group_by(fish_id, zone) %>%
  dplyr::filter(genotype %in% c("het", "hom")) %>% 
  mutate(aveTimeInZone = mean(time_in_zone), 
         zone = paste0("zone ", zone)) %>%
  dplyr::distinct(aveTimeInZone, .keep_all = TRUE) %>%
  ggplot(aes(x = treatment, y =aveTimeInZone)) +
  geom_quasirandom(aes(colour = genotype)
                  ) +
  geom_boxplot(aes(fill = genotype),
               outlier.shape = NA,
                alpha = 0.5) +
  facet_wrap(~zone, nrow = 1) +
  scale_y_log10() +
  scale_fill_viridis_d(end = 0.75) +
  scale_colour_viridis_d(end = 0.75) +
  easy_rotate_x_labels(angle = -45) +
  labs(y = "Average time spent in each zone (mins, log scale)", 
       title = "Average time spent in each zone")
```

## Statistical test

Here, I fit the average time spent in each zone of the maze to a linear mixed effect model. The `genotype x treatmetn x zone` effect is not significant, meaning there is not difference between the mutants and treatment groups. 

```{r}
fit <- read_csv("data/ymaze/processed_data/fecit/time_in_zone.csv") %>%
  mutate(fish_id = as.character(fish_id)) %>%
  dplyr::select(fish_id, bin, zone, time_in_zone) %>%
  left_join(meta) %>%
  group_by(fish_id, zone) %>%
  dplyr::filter(genotype %in% c("het", "hom")) %>% 
  mutate(aveTimeInZone = mean(time_in_zone), 
         zone = paste0("zone ", zone)) %>%
  dplyr::distinct(aveTimeInZone, .keep_all = TRUE) %>%
  lmer(log(aveTimeInZone) ~ (genotype + treatment + zone)^3 + (1|`start time`),
    data = .)

Anova(fit) %>% 
  kable(caption = "linear mixed model: aveTimeInZone) ~ (genotype + treatment + zone)^3 + (1|`start time`)" ) %>%
  kableExtra::row_spec(row = 7, bold = TRUE) %>% 
  kable_styling(full_width = FALSE)
```

# check for handedness

Fontana et al. (https://doi.org/10.1007/s10071-019-01296-9, Matt Parker's group) showed that fish sometimes show a behavioural lateralisation (i.e. handedness). If fish show this, then they would perform less alternation tetragrams not due to working memory.

The plot below circles each of the L_R bias groups. 

```{r}
# make the LR bias object
LR_Bias <- final_data %>%
  dplyr::select(L, R, total_turns, fish_id) %>%
  group_by(fish_id) %>%
  mutate(L = sum(L),
         R = sum(R),
         total_turns = sum(total_turns),
         L_R_bias = case_when( #consider more than 60% of the time performing a left or right turn to be a bias
           L/total_turns > 0.6 ~ "Left",
           R/total_turns > 0.6 ~ "Right",
           TRUE ~ "Neither"
         )) %>%
  dplyr::select(fish_id, L_R_bias) %>%
  unique() %>%
  mutate(L_R_bias = factor(L_R_bias,
                           levels = c("Neither", "Left", "Right"))
  )
```


```{r}
ggarrange(
final_data %>%
  left_join(LR_Bias) %>%
  group_by(fish_id) %>%
  mutate(L = sum(L),
         R = sum(R),
         total_turns = sum(total_turns)
  ) %>%
  ggplot(aes(L, R)) +
  geom_point(aes(shape = L_R_bias, colour = genotype),
             size = 4) +
  geom_mark_ellipse(aes(fill = L_R_bias, label = L_R_bias),
                 alpha = 0.2, 
                 con.cap = 0)+
  labs(title = "By genotype") +
  scale_color_viridis_d(end = 0.75) +
  theme(legend.position = "bottom", 
        aspect.ratio = 1) +
  scale_x_continuous(limits = c(-10, 650)) +
  scale_y_continuous(limits = c(-10, 650)) +
  scale_fill_viridis_d(option = "plasma"), 

final_data %>%
  left_join(LR_Bias) %>%
  group_by(fish_id) %>%
  mutate(L = sum(L),
         R = sum(R),
         total_turns = sum(total_turns)
  ) %>%
  ggplot(aes(L, R)) +
  geom_point(aes(shape = L_R_bias, colour = treatment),
             size = 4) +
  geom_mark_ellipse(aes(fill = L_R_bias, label = L_R_bias),
                 alpha = 0.2, 
                  con.cap = 0)+
  labs(title = "By treatment") +
  scale_color_viridis_d(end = 0.75) +
  scale_fill_viridis_d(option = "plasma") +
    scale_x_continuous(limits = c(-10, 650)) +
  scale_y_continuous(limits = c(-10, 650)) +
  theme(legend.position = "bottom", 
        aspect.ratio = 1)
)
```

# test for changes to alternation
Finally, I will now test for alternation changes using a generalised linear mixed effect model (beta-binomial distribution). We use this because it is count data which is over-dispersed, and fixed and random effects are included. The link function is logit.

No significant effects are observed on working memory. Only the `L_R_bias` term is significant, this is to be expected, as they would probably be performing more repitiions. 

The `bin:genotype` effect is close (ish) to statistical significance. Meaning that this might have something there. but this does not take into account the treatment group.   

Note that I have ignored the effect of sex here. We have never really seen a sex effect in all of our ymaze analyses. And if i were to include this effect, we would only have 1 and sometimes no fish within each genotype and sex. 

```{r}
glm <-
  final_data %>%
  left_join(LR_Bias) %>%
  mutate(
    non_alts = total_turns - alts,
    bin = as.factor(bin)
  ) %>%
  glmmTMB(
    cbind(alts, non_alts) ~ (bin + genotype + treatment)^3 + L_R_bias + (1|`start time`) + (1|fish_id),
    family = betabinomial(),
    data = .
  )

Anova(glm) %>%
  as.data.frame() %>%
  dplyr::rename(pval = `Pr(>Chisq)`) %>%
  kable() %>%
  kable_styling(full_width = FALSE) %>% 
  row_spec(row = 4, bold = TRUE)
```

## Vis 
### bin x genotype x treatment

```{r}
print(emmeans(glm, ~ genotype * treatment * bin), type = "response") %>% 
  as_tibble() %>% 
  mutate(binforvis = case_when(
    bin == 1 ~ "0-10 mins", 
    bin == 2 ~ "10-20 mins", 
    bin == 3 ~ "20-30 mins", 
    bin == 4 ~ "30-40 mins", 
    bin == 5 ~ "40-50 mins", 
    bin == 6 ~ "50-60 mins"
    )) %>% 
  ggplot(aes(x = binforvis, y = prob, colour = genotype)) +
  geom_col(aes(fill =genotype), 
           alpha = 0.5,
           width = 0.25,
           position = position_dodge()) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                width = 0.125,
                size = 1,
                position = position_dodge(width = 0.25)) +
  facet_wrap(~treatment, ncol = 1) +
  theme(axis.text.x = element_text(hjust = 1,
                               vjust = 1,
                               angle = 45), 
        legend.position = "bottom") +
  scale_color_viridis_d(end = 0.8, option = "inferno") +
  scale_fill_viridis_d(end = 0.8, option = "inferno") +
  labs(y = "Estimated probability of alternation\n(LRLR + RLRL)",
       x = "Time interval", 
       title = "GLM predicted probability of zebrafish performing an alternation\ntetragram due to genotype and treatment", 
       subtitle = "No significant differences"
       )
```

### L or R bias
```{r}
print(emmeans(glm, specs = "L_R_bias"), type = "response") %>%
  as_tibble() %>%
  ggplot(aes(L_R_bias, prob, colour = L_R_bias)) +
  geom_col(aes(fill = L_R_bias),
           alpha = 0.5,
           position = position_dodge(width = 0.5)) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.25,
    size = 1,
    position = position_dodge(width = 0.25)
    ) +
  ylab("Estimated probability of alternation") +
  xlab("Time interval") +
  theme(
    axis.text.x = element_text(hjust = 1,
                               vjust = 1,
                               angle = 45)
    ) +
  scale_y_continuous(limits = c(0,0.3)) +
  scale_color_viridis_d(end = 0.8, option = "viridis") +
  scale_fill_viridis_d(end = 0.8, option = "viridis") +
  ggtitle("GLM predicted probability of zebrafish performing an alternation\ntetragram due to having a LR bias",
          subtitle = "Effect of L or R Bias p = 0.002\naveraged out over genotypes, treatments and bins")
```

# test for changes in repetitions

Repetitions are a sign of stress. Matt Parker says he sees increased repetitions when fish are pre-treated with a chemogenic stressor. 

I fitted the over-bins data to another generalised linear mixed effect model (beta-binomial distribution), this time compariing the reps vs non-reps (LLLL + RRRR). 

Only the LR Bias term is significant. This is also expected. 

```{r}
glm_reps <- final_data %>%
  left_join(LR_Bias) %>%
  mutate(
    non_reps = total_turns - reps,
    bin = as.factor(bin)
  ) %>%
  glmmTMB(
    cbind(reps, non_reps) ~ (bin + genotype + treatment)^3 + L_R_bias + (1|`start time`) + (1|fish_id),
    family = betabinomial(),
    data = .
  )

glm_reps %>%
  Anova() %>%
  dplyr::rename(pval = `Pr(>Chisq)`) %>%
  kable() %>%
  kable_styling(full_width = FALSE) %>% 
  row_spec(row = 4, bold = TRUE)
```

## effect of genotype x treatment x bin

```{r}
print(emmeans(glm_reps, ~ genotype * treatment * bin), type = "response") %>% 
  as_tibble() %>% 
  mutate(binforvis = case_when(
    bin == 1 ~ "0-10 mins", 
    bin == 2 ~ "10-20 mins", 
    bin == 3 ~ "20-30 mins", 
    bin == 4 ~ "30-40 mins", 
    bin == 5 ~ "40-50 mins", 
    bin == 6 ~ "50-60 mins"
    )) %>% 
  ggplot(aes(x = binforvis, y = prob, colour = genotype)) +
  geom_col(aes(fill =genotype), 
           alpha = 0.5,
           width = 0.25,
           position = position_dodge()) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                width = 0.125,
                size = 1,
                position = position_dodge(width = 0.25)) +
  facet_wrap(~treatment, ncol = 1) +
  theme(axis.text.x = element_text(hjust = 1,
                               vjust = 1,
                               angle = 45), 
        legend.position = "bottom") +
  scale_color_viridis_d(end = 0.8, option = "inferno") +
  scale_fill_viridis_d(end = 0.8, option = "inferno") +
  labs(y = "Estimated probability of alternation\n(LRLR + RLRL)",
       x = "Time interval", 
       title = "GLM predicted probability of zebrafish performing an repetition\ntetragram due to genotype and treatment", 
       subtitle = "No significant differences"
       )
```

## L or R bias
```{r}
print(emmeans(glm_reps, specs = "L_R_bias"), type = "response") %>%
  as_tibble() %>%
  ggplot(aes(L_R_bias, prob, colour = L_R_bias)) +
  geom_col(aes(fill = L_R_bias),
           alpha = 0.5,
           position = position_dodge(width = 0.5)) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.25,
    size = 1,
    position = position_dodge(width = 0.25)
    ) +
  ylab("Estimated probability of alternation") +
  xlab("Time interval") +
  theme(
    axis.text.x = element_text(hjust = 1,
                               vjust = 1,
                               angle = 45)
    ) +
  scale_y_continuous(limits = c(0,0.4)) +
  scale_color_viridis_d(end = 0.8, option = "viridis") +
  scale_fill_viridis_d(end = 0.8, option = "viridis") +
  ggtitle("GLM predicted probability of zebrafish performing an repetition\ntetragram due to having a LR bias",
          subtitle = "Effect of L or R Bias p = 0.000008\naveraged out over genotypes, treatments and bins")
```

# Conclusion

