---
title: "FeDex_irf7"
author: "Karissa Barthelson"
date: "2022-10-31"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  autodep = TRUE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center", 
  out.width ="100%", 
  out.height = "100%"
)
```

```{r}
library(tidyverse)
library(readxl)
library(here)
library(magrittr)
library(scales)
library(readxl)
library(ggpubr)
library(ggeasy)
library(ggrepel)
library(ggfortify)
library(ggbeeswarm)
library(ggforce)
library(kableExtra)
library(emmeans)
library(performance)
library(see) # performance dependency
library(car)

# set the default theme for ggplot as theme_bw
theme_set(theme_bw())
```

# Introduction

Here, we are assessing the level of *irf7* transcripts in the brains of 6 month old zebrafish. These zebrafish are either heterozygous (unaffected) or homozygous (MPS-IIIB) for the A603fs mutation in *naglu*. They were treated with either vehicle (0.85% saline), 0.5 μg iron-dextran or 1 μg of iron-dextran . Then, they were allowed to recover alone in a pair mate tank for either 7 or 14 days. Here, we will determine whether there are significant changes to the expression of *irf7* due to *naglu* genotype and treatment. 

irf7 encodes a gene which is indicative of inflammation. We expect it to reduce after iron supplementation. 

```{r}
# read in daa
data <- read_xlsx("data/2022-10-19_201749_irf7_experimental_plate_1_384well.xlsx", 
                  sheet = "Results", skip = 40) 

meta <- read_csv("data/metadata.csv") %>% 
  mutate(fish_id = as.character(fish_id), 
         treatment = factor(treatment, 
                            levels = c("saline", 
                                       "0.5 ug iron-dextran", 
                                       "1 ug iron-dextran"))
  )

data %<>%  
  dplyr::rename("fish_id" = "Sample Name") %>% 
  left_join(meta) %>% 
  dplyr::select(colnames(meta), CT, everything()) %>% # put the metadata first for easy viewing
  mutate(CT = as.numeric(CT))
```

# Summary of samples

The plot below suymmarises the metadata of the fish we tested here. 

```{r}
meta %>% 
  group_by(genotype, sex, treatment, dpi) %>% 
  mutate(n = n()) %>% 
  ggplot(aes(x = genotype)) +
  geom_bar(aes(fill = treatment), 
           position = "dodge", 
           colour = "black") +
  facet_wrap(~dpi+sex, nrow = 1) +
  scale_fill_viridis_d() + 
  theme(legend.position = "bottom")+
  labs(title = "Number of fish per genotype, sex, treatment and dpi in the qPCR analysis",
       y = "Number of fish", 
       x = "")
```

# Check technical replicates of actin (housekeeping gene)

The following plot below shows the technical replicates of the actin gene. No problematic techincal replicates (well done Ewan :))

The no template control (NTC), has very high varation and CT value. This is normal though. gh. 

```{r}
# plot technical replicates for actin by plate. 
data %>% 
  dplyr::filter(fish_id %in% c("NTC", seq(1:100))) %>% # omit those dilution PCRs that ewan was playing with
  dplyr::filter(`Target Name` == "actin") %>% 
  unique %>% 
  mutate(fish_id = factor(fish_id, levels = c("NTC", seq(1:100)))) %>% # put in order
  ggplot(aes(x = fish_id, y = CT)) + 
  geom_point() +
  geom_boxplot(fill = NA) +
  geom_label_repel(aes(label = `Well Position`), # only label the variable fish
                   data = . %>% 
                       dplyr::filter(`Ct SD` > 1)) +
  ggtitle("actin technical replicates", 
          subtitle = "Well position is  labelled if the standard deviation of the tech reps is > 1")
```


# Check technical replicates of irf7 

The following plot below shows the technical replicates of the irf7 gene. Again,no problematic technical replicates (apart from the NTC, which is normal and maybe a little bit for fish 1 and 5). However, there appears to be more variability for this gene. 


```{r}
# plot technical replicates for tfr1a by plate. 
data %>% 
  dplyr::filter(fish_id %in% c("NTC", seq(1:100))) %>% # omit those dilution PCRs that ewan was playing with
  dplyr::filter(`Target Name` == "irf7") %>% 
  unique %>% 
  mutate(fish_id = factor(fish_id, levels = c("NTC", seq(1:100)))) %>% # put in order
  ggplot(aes(x = fish_id, y = CT)) + 
  geom_point() +
  geom_boxplot(fill = NA) +
  geom_label_repel(aes(label = `Well Position`), # only label the variable fish
                   data = . %>% 
                       dplyr::filter(`Ct SD` > 1)) +
  ggtitle("tfr1a technical replicates", 
          subtitle = "fish are labelled if the standard deviation of the tech reps is > 1")
```


# Plot the results 

Expression of irf7 appears to increase with higher amounts of iron. 

```{r}
data %>% 
  unique() %>% 
  dplyr::select(colnames(meta), `Target Name`, CT) %>% 
  dplyr::filter(genotype %in% c("het", "hom")) %>% # only select fish of interest
  group_by(fish_id, `Target Name`) %>% 
  mutate(aveCT = mean(CT)) %>% 
  dplyr::select(-CT) %>% 
  unique %>% 
  spread(`Target Name`, aveCT) %>% 
  mutate(dct = irf7 - actin ) %>% 
    ggplot(aes(x = genotype, y = dct, fill = genotype)) +
  geom_boxplot(aes(fill = genotype), 
               alpha = 0.5,
               outlier.shape = NA) +
  geom_point(size =3, 
             aes(shape = sex)) +
  scale_fill_viridis_d(end = 0.5) +
  facet_wrap(~dpi+treatment, nrow = 1) +
  theme(legend.position = "bottom") +
  scale_y_continuous(limits = c(0,9)) +
  ggtitle("boxplots of tfr1a", 
          subtitle = "expression of tfr1a relative to beta actin") 

data %>% 
  unique() %>% 
  dplyr::select(colnames(meta), `Target Name`, CT) %>% 
  dplyr::filter(genotype %in% c("het", "hom")) %>% # only select fish of interest
  group_by(fish_id, `Target Name`) %>% 
  mutate(aveCT = mean(CT)) %>% 
  dplyr::select(-CT) %>% 
  unique %>% 
  spread(`Target Name`, aveCT) %>% 
  mutate(dct = irf7 - actin, 
         dpi = case_when(
           dpi == "7" ~ "7 dpi",
           dpi == "14" ~ "14 dpi"
         ) %>% 
           factor(levels = c("7 dpi", "14 dpi")), 
         sex = factor(sex, levels = c("female", "male"))
         ) %>% 
  ggplot(aes(x = treatment, y = dct, fill = genotype, colour = genotype)) +
  stat_summary(fun=mean, # add mean bar plot
               alpha = 0.5,
               geom="bar", 
               position = "dodge") +
  geom_point(
             alpha=0.9, 
             position = position_jitterdodge(jitter.width = .1, dodge.width = 1),
             size = 3
             
  ) +
  facet_wrap(~dpi, nrow = 1) +
  scale_fill_viridis_d(end = 0.5) +
  scale_colour_viridis_d(end = 0.5) +
    ggtitle("barplots showig mean expression of irf7", 
          subtitle = "expression of irf7 relative to beta actin") 
```


# Statistical test

## Check model assumptions

I fit a linear model and checked assumptions using the `performance` package. Overall, they dont look too bad. Some colinearity is observed. Unsure if this is an issue as of yet. Is it because we have very low n per genotype, treatment and dpi?

```{r}
# fit the lm (interactive model), 
fit <- 
  data %>% 
  unique() %>% 
  dplyr::select(colnames(meta), `Target Name`, CT) %>% 
  dplyr::filter(genotype %in% c("het", "hom")) %>% # only select fish of interest
  group_by(fish_id, `Target Name`) %>% 
  mutate(aveCT = mean(CT)) %>% 
  dplyr::select(-CT) %>% 
  unique %>% 
  spread(`Target Name`, aveCT) %>% 
  mutate(dct = irf7 - actin )   %>% 
  lm(dct ~ genotype*treatment*dpi, 
     data = . )
```


```{r, fig.height=15}
# check assumptions
# overall not too bad, collinearity is observed. 
check_model(fit)
```

Nevertheless, I checked the significnace of the effects and their interactions using type II Anova. While the effect of `genotype * treatment * dpi` is not significant, this is likely due to being very underpowered. 

```{r}
# test for significance of effects 
Anova(fit) %>%
  dplyr::rename(pval = "Pr(>F)") %>% 
  kable(caption = "Anova Table (Type II tests)") %>% 
  kable_styling(full_width = F)
```




